
---
title: "epidemia"
output: 
  github_document:
    fig_width: 10
    fig_height: 6
bibliography: inst/REFERENCES.bib
vignette: >
  %\VignetteIndexEntry{README}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

<style type="text/css">
body{ /* Normal */
font-size: 16px;
}
</style>

**epidemia** is an R package allowing flexible 
specification and fitting of Bayesian models for epidemics in the style of 
@Flaxman2020. The package leverages R's formula interface to paramaterize the time-varying reproduction rate as a function of covariates. Multiple populations can be modeled simultaneously using hierarchical models. The design of the package has been inspired by, and has borrowed from, [rstanarm](https://mc-stan.org/rstanarm/) [@rstanarm]. epidemia uses [rstan](https://mc-stan.org/rstan/) [@rstan] as the backend for fitting models.

This is an early beta release of the package. There will be regular updates with additional features and more extensive testing. Any feedback is greatly appreciated - in particular if you find bugs, find the documentation unclear, or have feature requests, please report them [here](https://github.com/ImperialCollegeLondon/epidemia/issues).

## Installation

The repository is currently private. If you are interested in beta testing, please [contact us](mailto:james.scott15@imperial.ac.uk). Once you have access to the repository, clone it and install from source. We recommend avoiding installation of any vignettes as these are computationally demanding, and are best viewed online.

## Getting Started

The best way to get started is to read the [vignettes](articles/index.html).

* [Flexible Epidemic Modeling with epidemia](articles/introduction.html) is the main vignette, introducing the model framework and the primary model fitting function `epim`.
* [Incidence Only](articles/IncidenceOnly.html) gives examples of fitting models using only incidence data and a discrete serial interval.
* [Time-Dependent Modeling](articles/TimeDependentR.html) demonstrate how to model weekly changes in the reproduction number as a random walk.

## Usage

```{r exampleUsage}
library(epidemia)
library(xfun)
options(mc.cores=parallel::detectCores())

data(EuropeCovid)
# Collect args for epim
args <- EuropeCovid
args$algorithm <- "sampling"
args$group_subset <- c("Germany", "United_Kingdom")
args$sampling_args <- list(iter=1e3,control=list(adapt_delta=0.95,max_treedepth=15),seed=12345)
args$formula <- R(country, date) ~ schools_universities + self_isolating_if_ill +
  public_events + lockdown + social_distancing_encouraged
args$prior <- shifted_gamma(shape = 1/6, scale = 1, shift = -log(1.05)/6)
fit <- xfun::cache_rds({do.call("epim", args)}, hash=args)
```

```{r examplePlots}
# Inspect Rt
plot_rt(fit, group = "United_Kingdom", levels = c(25,50,75,95))
# And deaths
plot_obs(fit, type = "deaths", group = "United_Kingdom", levels = c(25,50,75,95))
```

## References