---
title: "Modelling of R with Case data only"
output: 
  html_document
vignette: >
  %\VignetteIndexEntry{Modelling of R  with only Incidence data}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

This vignette describes how changes in the reproduction number can be modelled with just case data and 
with a discrete serial interval.

## Flu 1918

Data from the 1918 Flu pandemic is used as an example. This data is provided in the R package @EpiEstim.

```{r Flu1918comp, message=FALSE}
library(epidemia)
library(EpiEstim)
library(plotly)
data("Flu1918")
options(mc.cores = parallel::detectCores())
print(Flu1918)
flu <- Flu1918
```

First, we form the `data` argument to `epim`. Recall that this must store all observations and covariates needed to fit the model.
In our case, this is easy because there are no covariates, and there is just one observation type - case data.

```{r}
data <- data.frame(
  country = "A",
  cases = c(NA, flu$incidence), # pad start
  fludate = as.Date("1918-01-01") + seq(0, along.with=c(NA,flu$incidence)) 
)

# needed for weekly random walk
data$week <- format(data$fludate, "%V")
```

A week column as been added to `data` because we will parameterise the reproduction number as a weekly random walk. 
This is done as follows.
```{r}
# model the rep number as weekly random walk (no covariates)
rt <- epirt(
  formula = R(country, fludate) ~ 0 + rw(time=week, prior_scale=0.1),
  r0=3 # prior expected reproduction number
)
```

We will also need to specify a model for how the observations (daily cases) are observed from the latent infection series.
This is done through the `epiobs` function.
```{r}
# model observed cases as a proportion of infections constant over time
cases <- epiobs(
  formula = cases(country, fludate) ~ 1,
  prior_intercept = rstanarm::normal(location=1, scale=0.01),
  link = "identity",
  lag = rep(.25,4)
)
```
In this case, we assume that everybody infected is recorded as a case, and that this happens with equal probability in the four 
days subsequent to infection.

We will use the NUTs sampler to fit the model. It is useful to explicitly control the parameters for sampling.
```{r}
# use NUTs with given parameters
sampling_args <- list(
  iter=1000,
  control=list(adapt_delta=0.95,max_treedepth=15),
  seed=12345
)
```
We are ready to fit the model. This is done through a call to `epim`.
```{r Flu1918fit, cache=TRUE}
# fit the model with various model params
fm <- epim(
  rt = rt,
  obs = list(cases),
  data = data,
  pops = data.frame(country="A", pop=1e6),
  si = flu$si_distr,
  seed_days = 6,
  prior_tau = rstanarm::exponential(rate=4),
  algorithm = "sampling",
  sampling_args = sampling_args
)
```

The resulting fit can be examined using various plotting functions provided by **epidemia**.
```{r, warning=FALSE, fig_width=15}
subplot(
plot_rt(fm, plotly=T, smooth=7),
plot_infections(fm, plotly=T),
plot_obs(fm, type = "cases", plotly=T),
plot_infectious(fm, plotly=T),
shareX = T,
nrows=2,
margin=0.04
)
```


