---
title: "Implementation"
output:
  bookdown::html_document2:
    theme: flatly
    number_sections: true
link-citations: yes
bibliography: ../inst/REFERENCES.bib
pkgdown:
  as_is: true
---

<style>
.contents .page-header {
    display: none;
} 
</style>


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(kableExtra)
```

<div class = "article_container">

# Model Implementation {#sec:modelimplementation}
<hr>

Here we give a high-level overview of the workflow required for defining and fitting a model with **epidemia**. The primary model fitting function is `epim()`. This takes a model description and additional arguments relating to the fitting algorithm, and proceeds to fit the model using a precompiled **Stan** program. This is similar to the workflow for fitting Bayesian regression models 
with **rstanarm**. A key difference, however, is that the models fit by **epidemia** are generally complex, and are therefore inherently more difficult to specify. We simplify this process by taking a modular approach; models are defined through three distinct parts: transmission, infections and observations. These components of the model are defined with the functions `epirt()`, `epiinf()` and 
`epiobs()` respectively.

The package contains an example data set `EuropeCovid` which contains data 
on daily death counts from Covid-19 in 11 European Countries from February through May 2020, and a set 
of binary indicators of non-pharmaceutical interventions. This is 
used as an example throughout.

```{r model-imp-load-packages, message=FALSE}
library(dplyr)
library(epidemia)
library(rstanarm)
data("EuropeCovid")
```

We begin by describing `epim()` in more detail, and then proceed to discuss the three modeling functions. 

## Model Fitting  {#sec:fitting}
<hr>

`epim()` is the only model fitting function in **epidemia**. It has arguments `rt`, `inf`, and `obs` which 
expect a description of the transmission model, infection model and all observational models respectively. 
Together, these fully define the joint distribution of data and parameters. Each of these model components are 
described in terms of variables that are expected to live in a single data frame, `data`. This data frame 
must be compatible with the model components, in the sense that *it holds all variables defined in these 
models*. For our example, these variables are the following. 
```{r model-imp-data-vars}
data <- EuropeCovid$data
colnames(data)
```
The `data` argument is described in more detail in Section \@ref(sec:data).

In addition to taking a model description and a data frame, 
`epim()` has various additional arguments which specify how the model should be
fit. If `algorithm = "sampling"` then the model 
will be fit using **Stan**'s adaptive Hamiltonian Monte Carlo sampler [@hoffman_2014]. This 
is done internally by calling `sampling()` from **rstan**. If instead this is `"meanfield"` or `"fullrank"`, then **Stan**'s Variational Bayes 
algorithms [@Kucukelbir_2015; @Kucukelbir_2017] are employed by calling `vb()` from **rstan**. Any unnamed 
arguments in the call to `epim()` are passed directly onto the **rstan**
sampling function. `epim()` returns a fitted model object of class 
`epimodel`, which contains posterior samples from the model along with other 
useful objects.

In general, Hamiltonian Monte Carlo should be used for final inference. Nonetheless, this is often computationally demanding, and Variational Bayes can often be 
used fruitful for quickly iterating models. All arguments for  `epim()` are described in Table 
\@ref(tab:model-imp-epim-args).


<div class="table">
```{r model-imp-epim-args, echo=FALSE}
arguments <- c(
    "`rt`",
    "`inf`",
    "`obs`",
    "`data`",
    "`algorithm`",
    "`group_subset`",
    "`prior_PD`",
    "..."
)

description <- c(
    paste0("An object of class `epirt`, resulting from a call to `epirt()`. This defines the model for ",
    "time-varying reproduction numbers $R_t$."),
    paste0("An object of class `epiinf`, resulting from a call to `epiinf()`. This entirely defines ",
    "the model for infections $i_t$."),
    paste0("Either an object of class `epiobs`, or a list of such objects. Each of these define a model for an observation vector in `data`, and result from a call to `epiobs()`.",
       " Each element of the list defines a model for an observed variable."),
    paste0("A dataframe with all data required for fitting the model. This includes all observations and covariates specified in the model."),
    paste0("One of `\"sampling\"`, `\"meanfield\"` or `\"fullrank\"`. This determines the **rstan** sampling ",
    "function to use for fitting the model. `\"sampling\"` corresponds to HMC, while ",
    "`\"meanfield\"` and `\"fullrank\"` are Variational Bayes algorithms."),
    "If specified, a character vector naming a subset of groups/populations to include in the model.",
    paste0("If `TRUE`, parameters are sampled from their prior distributions. This is ",
    "useful for prior predictive checks. Defaults to `FALSE`."),
    paste0("Additional arguments to pass to the **rstan** function used to fit the model. If ", 
           "`algorithm = \"sampling\"`, then this function is `sampling()`. Otherwise `vb()` is used.")
)
df <- data.frame(Argument = arguments, Description = description)
knitr::kable(df, caption = "Formal arguments for the model fitting function `epim()`. The first three arguments listed below define the model to be fitted.", booktabs=TRUE, escape = FALSE) %>%
  kable_paper("hover") %>%
  column_spec(1, bold = T) 
```
</div>

## Transmission  {#sec:imp_transmission}
<hr>

`epirt()` defines the model for time-varying reproduction numbers, which 
was described in Section 1.4 of the model description [article](model-description.html). Recall that these are modeled as a transformed linear predictor. `epirt()` has a `formula` argument which defines the linear predictor $\eta$, an argument `link` defining the link function `g`, and additional arguments to specify priors on parameters making up $\eta$.

A general **R** formula gives a symbolic description of a model. It takes the form 
`y ~ model`, where `y` is the response and `model` is a collection of terms 
separated by the `+` operator. `model` fully defines a linear predictor used to 
predict `y`. In this case, the "response" being modeled are reproduction numbers which are
unobserved. `epirt()` therefore requires that 
the left hand side of the formula takes the form `R(group, date)`, where `group` 
and `date` refer to variables representing the modeled populations and dates respectively. 
The right hand side can consist of fixed effects, random effects, and 
autocorrelation terms. For our example, a viable call to `epirt()` is the 
following.
```{r model-imp-epirt-example}
rt <- epirt(formula = R(country, date) ~ 1 + lockdown + public_events,
            link = scaled_logit(7))
```
Here, two fixed effects are included which represent the 
effects of implementing lockdown and banning public events. These effects are assumed constant across 
countries. They could alternatively be partially 
pooled by using the term `(lockdown + public_events | country)`. 
For information on how to interpret such terms, please see [partial pooling](partial-pooling.html). Using `link = scaled_logit(7)` lets the link function be the 
scaled logit link described by Equation (1.7) in the model description article, 
where $K = 7$ is the maximum possible value for reproduction numbers.
For simplicity, we have omitted any prior arguments, however these 
should generally be specified explicitly. Please see 
[prior](priors.html) for detailed information on how to use priors. All arguments 
for `epirt()` are listed in Table \@ref(tab:model-imp-epim-args).


<div class="table">
```{r model-imp-epirt-args, echo=FALSE}
arguments <- c(
    "`formula`", 
    "`link`", 
    "`center`", 
    "`prior`", 
    "`prior_intercept`", 
    "`prior\\_covariance`",
    "`...`")

description <- c(
    paste0(
    "An object of class `formula` which determines the linear predictor $\\eta$ for $R$. ", 
    "The left hand side must take the form `R(group, date)`, where `group` must be ",
    "a factor vector indicating group membership (i.e. country, state, age cohort), and `date` must ",
    "be a vector of class `Date`. This is syntactic sugar for the reproduction number in the given group at the give date."),
    "The link function $g$. Can be `\"log\"`, `\"identity\"` or a call to `scaled_logit()`. Defaults to `\"log\"`.",
    paste0("If `TRUE`, covariates specified in `formula` are centered to have mean zero. All priors should then be ",
    "interpreted as priors on the centered covariates."),
    paste0("Same as in `stan_glm()` from **rstanarm**. Defines the prior on fixed effects $\\beta$. ", 
    "Priors provided by **rstanarm** can be used, and additionally `shifted_gamma`. **Note**: ",
"if `autoscale = TRUE` in the call to the prior function, then automatic rescaling takes place."),
    "Same as in `stan_glm()` from **rstanarm**. Prior for the regression intercept $\\beta_0$ (if it exists).",
    paste0("Same as in `stan_glmer()` from **rstanarm**. Defines the prior on the covariance matrix ",
    "$\\Sigma$. Only use if the `formula` has one or more terms  of the form `(x | y)`, ",
    "in which case there are parameters to partially pool, i.e. $b$ has positive length."),
  paste0("Additional arguments to pass to `model.frame()` from **stats**.")
)

df <- data.frame(Argument = arguments, Description = description)
knitr::kable(df, caption = "Formal arguments for `epirt()`, which defines the model for $R_t$.", booktabs=TRUE, escape=FALSE) %>%
  kable_paper("hover") %>%
  column_spec(1, bold = T)
```
</div>

## Infections  {#sec:imp_infections}
<hr>

The infection model is represented by `epiinf()`. In the most basic version, this defines the distribution of the generation time of the disease, the number of days for which to seed infections, and the prior distribution on seeded infections. These three parameters are 
controlled by the arguments `gen`, `seed_days` and `prior_seeds` respectively. 
A possible model is the following.
```{r model-imp-epiinf-example}
inf <- epiinf(gen = EuropeCovid$si, seed_days = 6L, 
              prior_seeds = hexp(exponential(0.02)))
```
`EuropeCovid$si` is a numeric vector representing the distribution for the serial interval of Covid-19. There is an implicit assumption that the generation time can be approximated well by the serial interval. Seeds are modeled hierarchically, and are described by (1.4) and (1.5) in the model description. $\tau$ has been assigned an exponential prior with a mean of 50. Seeded infections are assumed to occur over a period of 6 days.

`epiinf()` has additional arguments that allow the user to extend the basic 
model. Using `latent = TRUE` replaces the renewal equation with 
Equation (1.8). Daily infections are then treated as latent parameters that are
sampled along with other parameters. The `family` argument specifies 
the distribution $p(i'_t, d)$, while `prior_aux` defines the prior on the 
coefficient of dispersion $d$.

Recall from the model description that the infection process may be 
modified to explicitly account for changes in infection rates as the remaining 
susceptible population is depleted. In particular, the adjustment ensures that cumulative infections 
never breaches the population size. It can be employed by 
setting `pop_adjust = TRUE` and using the 
`pop` argument to point towards a  static variable in the data frame giving the population size. All argument to `epiinf()` are 
described in Table \@ref(tab:model-imp-epiinf-args).

<div class="table">
```{r model-imp-epiinf-args, echo=FALSE}
arguments <- c(
    "`gen`",
    "`seed_days`",
    "`prior_seeds`",
    "`latent`",
    "`family`",
    "`prior_aux`",
    "`fixed_vtm`",
    "`pop_adjust`",
    "`pops`",
    "`prior_susc`",
    "`rm`",
    "`prior_rm_noise`"
)

description <- c(
    paste0("A numeric vector giving the probability mass function $g_k$ for the generation time of the disease (must be a probability vector)."),
    "An integer giving the number of days $v + 1$ for which to seed infections. Defaults to 6L.",
    "Prior distribution on the seed parameter $i$. Defaults to  `hexp(prior_aux = rstanarm::exponential(0.03))`.",
    "If `TRUE`, treat infections as latent parameters.",
    "Specifies the family for the infection distribution $p(i'_t, d)$. Only used if `latent = TRUE`, and defaults to `\"normal\"`.",
    "Prior on the auxiliary variable $d$ of $p(i'_t,d)$. This is either the variance-to-mean ratio or the coefficient of variation, depending on the value of `fixed_vtm`. Only used if `latent = TRUE`.",
    "If `TRUE`, then $p(i'_t, d)$ has a fixed variance-to-mean ratio, i.e. variance is $\\sigma^2 = d i'_t$; In this case, $d$ refers to the *variance-to-mean ratio*. Id `FALSE` then instead standard deviation is assumed proportional to the mean, in which case $d$ is the *coefficient of variation*. Only used if `latent = TRUE`.", 
    "If `TRUE`, applies the population adjustment to the infection process.",
    "A character vector giving the population variable. Only used if `pop_adjust = TRUE`.",
    "Prior on $S_{v-1} / P$, the initial susceptible population as a proportion of the population size. If `NULL`, this is assumed to be equal to 1 (i.e. everyone is initially susceptible). Otherwise, can be a call to `normal()` from **rstanarm**, which assigns a normal prior truncated to $[0,1]$. Only used if `pop_adjust = TRUE`.",
    "A character vector giving the variable corresponding to $v_t$, i.e. the proportion of $S_t$ to remove at time $t$. Only used if `pop_adjust = TRUE`.",
    "Prior on the parameter $\\xi$, which controls noise around $v_t$. If `NULL`, no noise is added. Only used if `pop_adjust = TRUE`."
)

df <- data.frame(Argument = arguments, Description = description)
knitr::kable(df, caption = "Formal arguments for `epiinf()`, which defines the infection model.", booktabs=TRUE, escape=FALSE) %>%
  kable_paper("hover") %>%
  column_spec(1, bold = T)
```
</div>


## Observations  {#sec:imp_observations}
<hr>

An observational model is defined by a call to `epiobs()`. In particular, 
this must also make explicit the model for the multipliers $\alpha_t$, and must 
also specify the coefficients $\pi_k$. `epiobs()` has a `formula` argument. The left hand 
side must indicate the observation vector to be modeled, while the 
right hand side defines a linear predictor for $\alpha_t$. The argument `i2o` plays a similar role to the `gen` argument in `epiinf()`, 
however it instead corresponds the vector $\pi$ in Equation (1.2).

Take for example the task of modeling daily `deaths`, which as we saw is 
a variable in `data`. A possible model is the following. 
```{r model-imp-epiobs-example}
deaths <- epiobs(formula = deaths ~ 1, i2o = EuropeCovid$inf2death, 
                 link = scaled_logit(0.02))

```
Here $\alpha_t$ corresponds to the infection fatality rate (IFR), and is modeled as an 
intercept transformed by the scaled-logit link. This implies that the IFR is constant over time and its value lies somewhere between 0\% and 2\%. If the prior on 
the intercept (specified by the `prior_intercept` argument) is chosen to be 
symmetric around zero, then the prior mean for the IFR is 1\%. `EuropeCovid$inf2death` is a numeric simplex vector that gives the same delay distribution as used in 
@Flaxman2020. This is a density function for a discretized mixture of Gamma 
random variables.

Additional arguments include `family`, which specifies the sampling distribution 
$p(y_t, \phi)$. There are also arguments allowing the user to control 
prior distributions for effects in the linear predictor for $\alpha_t$, and the prior on the 
auxiliary variable $\phi$. All arguments to `epiobs()` are shown in Table 
\@ref(tab:model-imp-epiobs-args).


<div class="table">
```{r model-imp-epiobs-args, echo = FALSE}
arguments <- c(
    "`formula`",
    "`i2o`",
    "`family`",
    "`link`",
    "`center`,\n `prior`,\n `prior_intercept`",
    "`prior_aux`",
    "`...`"
)

description <- c(
    paste0("An object of class `formula` which determines the linear predictor for the ",
    "ascertainment rate. The left hand side must define the ",
    "response that is being modeled (i.e. the actual observations, not the latent ascertainments)"),
    paste0("A numeric (probability) vector defining the probability mass function $\\pi_k$ of the ",
    "time from an infection to an observation."),
    paste0("A string representing the family of the sampling distribution $p(y_t,\\phi)$. Can be ",
    "one of `\"poisson\"`, `\"neg_binom\"`, `\"quasi_poisson\"`, `\"normal\"` or `\"log_normal\"`."),
    paste0("A string representing the link function used to transform the linear predictor. Can be ",
    "one of `\"logit\"`, `\"probit\"`, `\"cauchit\"`, `\"cloglog\"`, `\"identity\"`. Defaults to `\"logit\"`."),
    "same as in `epirt()`, described above.",
    paste0("The prior distribution for the auxiliary parameter $\\phi$, if it exists. ",
    "Only used if family is `\"neg_binom\"` (reciprocal dispersion), `\"quasi_poisson\"` (dispersion), ",
    "`\"normal\"` (standard deviation) or `\"log_normal\"` (sigma parameter)."),
    "Additional arguments for `model.frame()` from **stats**."
)
df <- data.frame(Argument = arguments, Description = description)
knitr::kable(df, caption = "Formal arguments for `epiobs()`. This defines a single observation model. Multiple such models can be used and passed to `epim()` in a list.", booktabs=TRUE, escape=FALSE) %>%
  kable_paper("hover") %>%
  column_spec(1, bold = T, width = "7em") 
```
</div>

## Data {#sec:data}
<hr>

Before fitting our first model in Section \@ref(sec:first-fit), we elaborate on 
the `data` argument to `epim()`. Recall that this must contain all variables used in the transmission and infection models, and in all observational models. For our example, 
`data` looks like 
```{r model-imp-covid-data}
head(data)
```
The columns `country` and `date` define the region and time period corresponding to each of the remaining 
variables. `epim()` assumes that the first seeding day (i.e. the start of the epidemic) in each region is the first date found in the data frame. The last data found for each region is the final data at which the epidemic is simulated. It is up to the user to appropriately choose these dates. For our example,
the first and last dates for each group can be seen as follows.
```{r model-imp-covid-data-dates}
dates <- summarise(data, start = min(date), end = max(date))
head(dates)
```
Here, the start dates have been heuristically chosen to be 30 days prior to 
observing 10 cumulative deaths in each country.

## A First Fit {#sec:first-fit}
<hr>

We are now ready to fit our first model. For this we return to the model 
fitting function `epim()`. The following command is used to instruct 
**epidemia** to run Markov chains in parallel, rather than sequentially, if 
multiple cores are detected.
```{r model-imp-parallel}
 options(mc.cores = parallel::detectCores())
```
Our call to `epim()` is as follows. We use `refresh = 0` 
to suppress printing output in this article, however, this should 
not generally be used as such output is useful. 

```{r model-imp-fit-model}
fm <- epim(rt = rt, inf = inf, obs = deaths, data = data, 
           group_subset = "France", algorithm = "sampling", iter = 1e3, 
           seed = 12345, refresh = 0)
```

The print method for `epimodel` objects prints summary statistics for 
model parameters. These are obtained from the sampled posterior distribution. 
Parameter are displayed 
according to which part of the model they belong to (transmission, observations,
infections). An estimate of the standard deviation, labeled `MAD_SD` is 
displayed. This is the median absolute deviation from the median, and is more 
robust than naive estimates of the standard deviation for long-tailed 
distributions. 

```{r model-imp-fit-print}
print(fm)
```

Alternatively, the summary method can be used. This gives quantiles of the 
posterior draws, and also displays some MCMC diagnostics.

```{r model-imp-fit-summary}
summary(fm)
```

# References
<hr>

</div>
