---
title: "Implementation"
output:
  bookdown::html_document2:
    theme: flatly
    number_sections: true
link-citations: yes
bibliography: ../inst/REFERENCES.bib
pkgdown:
    as_is: true
---


<style>
.col-md-9 {
    width: 80% !important;
    margin-left: auto;
    margin-right: auto;
    line-height: 2;
    text-align: justify;
}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


#  Overview {#sec:overview}

Here we give a high-level overview of the workflow required for defining and fitting a model with 
epidemia. The primary model fitting function is `epim()`. This takes a model description and 
additional arguments relating to the fitting algorithm, and proceeds to fit the model using a 
precompiled Stan program. This allows model fitting to begin immediately as opposed to requiring compilation each time 
`epim()` is called. 

This is similar to the workflow for fitting Bayesian regression models 
with rstanarm. A key difference, however, is that the models fit by epidemia are much more 
complex, and are therefore inherently more difficult to specify. epidemia aims to simplify this process 
by modularising the model definition into three distinct parts: transmission, 
infections and observations. These components of the model are defined with the functions `epirt()`, `epiinf()` and 
`epiobs()` respectively.

We begin by describing `epim()` in more detail, and then proceed to discuss the three modeling functions. 
epidemia contains an example dataset `EuropeCovid` which contains, among other things, data 
on daily death counts from Covid19 in 11 European Countries from Feb through March? This is 
used as a running example throughout this article.

```{r load-packages, message=FALSE}
library(epidemia)
library(rstanarm)
data("EuropeCovid")
```

# Model Fitting

`epim()` is the only model fitting function in epidemia. It has arguments `rt`, `inf`, and `obs` which 
expect a description of the transmission model, infection model and all observational models respectively. 
Together, these fully define the joint distribution of data and parameters. 

```{r epim-args, echo=FALSE}
arguments <- c(
    "`rt`",
    "`inf`",
    "`obs`",
    "`data`",
    "`algorithm`",
    "`sampling_args`",
    "`prior_PD`"
)

description <- c(
    paste0("An object of class `\"epirt\"`, resulting from a call to `epirt()`. This defines the model for ",
    "reproduction numbers $R$."),
    paste0("An object of class `\"epiinf\"`, resulting from a call to `epiinf()`. This entirely defines ",
    "the model for infections $i_t$.)"),
    paste0("A list of objects of class `epiobs`. Each element of the list defines a model for a given ",
    "observation vector in `data`. )"),
    paste0("A dataframe with all data required for fitting the model. This includes all observation ",
    "columns and predictors specified in the models for the reproduction number and ascertainment rates."),
    paste0("One of `\"sampling\"`, `\"meanfield\"` or `\"fullrank\"`. This determines the rstan sampling ",
    "function to use for fitting the model. `\"sampling\"` corresponds to an adaptive HMC sampler, while ",
    "`\"meanfield\"` and `\"fullrank\"` are both Variational Bayes algorithms."),
    paste0("An (optional) named list of parameters to pass to the rstan sampling function, ",
    "for example `rstan::sampling`."),
    paste0("Same as in `rstan::stan_glm`. If `TRUE`, samples parameters from their priors. This is ",
    "useful for prior predictive checks. Defaults to `FALSE`.")
)
df <- data.frame(Argument = arguments, Description = description)
knitr::kable(df, caption = "Formal arguments for `epim()`", booktabs=TRUE)
```

# Transmission

```{r epirt-args, echo=FALSE}
arguments <- c(
    "`formula`", 
    "`link`", 
    "`center`", 
    "`prior`", 
    "`prior_intercept`", 
    "`prior_covariance`")

description <- c(
    paste0(
    "An object of class `\"formula\"` which determines the linear predictor $\\eta$ for $R$. ", 
    "The left hand side must take the form `R(group_col, date_col)`, where `group_col` and `date_col` are ",
    "column names in the data frame passed as the `data` argument to `epirt()`. `group_col` must be ",
    "a factor vector indicating group membership (i.e. country, state, age cohort), and `date_col` must ",
    "be a vector of class `\"Date\"`. This is syntactic sugar for the reproduction number in the given group at the give date."),
    "The link function $g$. Can be `\"log\"`, `\"identity\"` or a call to `scaled_logit()`. Defaults to `log`.",
    paste0("If `TRUE`, covariates in `formula` are centered to have mean zero. All priors should then be ",
    "interpreted as priors on the centered covariates."),
    paste0("Same as in `rstanarm::stan_glm`.  Defines the prior on $\\beta$. ",
    "`rstanarm` provided [rstanarm priors](http://mc-stan.org/rstanarm/reference/priors.html), a `shifted_gamma` can be used. **Note** ",
"if `autoscale=TRUE` in the call to the prior function, then automatic rescaling takes place."),
    "Same as in `rstanarm::stan_glm`. Prior for the regression intercept $\\beta_0$ (if it exists).",
    paste0("Same as in `rstanarm::stan_glmer`. Defines the prior on the covariance matrix ",
    "$\\Sigma$. Only use if the `formula` has one or more terms  of the form `(x | y)`, ",
    "in which case there are parameters to partially pool, i.e. $b$ has positive length.")
)

df <- data.frame(Argument = arguments, Description = description)
knitr::kable(df, caption = "Formal arguments for `epirt()`", booktabs=TRUE)
```

# Infections

```{r epiinf-args, echo=FALSE}
arguments <- c(
    "`seed_days`",
    "`gen`",
    "`prior_tau`",
    "`latent`",
    "`family`",
    "`prior_aux`",
    "`pop_adjust`",
    "`susceptibles`"
)

description <- c(
    "An integer giving the number of seed days $v + 1$. Defaults to 6L.",
    paste0("A numeric vector giving the probability mass function $g_k$ of the generation distribution. ",
    "Must be a simplex vector, i.e. nonnegative and summing to 1."),
    paste0("Prior distribution for the hyperparameter $\\tau$, which is the mean of the prior ",
    "distribution for infection seeding. Defaults to `rstanarm::exponential(0.03)`."),
    "If `TRUE`, treat infections as latent parameters using the extensions described in Section XXX. ",
    "Specifies the family for $p(i'_t, d)$. Only used if `latent = TRUE`, and currently restricted to `log-normal`.",
    "Prior on the coefficient of variation $d$.",
    "If `TRUE`, applies a population adjustment to the infection process $i_t$. Defaults to `FALSE`.",
    paste0("A character vector giving the name of the column in the dataframe passed as the `data` argument of `epim`, that ",
    "corresponds to the susceptible population over time. Only used if `pop_adjust=TRUE`.")
)

df <- data.frame(Argument = arguments, Description = description)
knitr::kable(df, caption = "Formal arguments for `epiinf()`", booktabs=TRUE)
```

# Observations

```{r epiobs-args, echo = FALSE}
arguments <- c(
    "`formula`",
    "`i2o`",
    "`family`",
    "`link`",
    "`center`, `prior`, `prior_intercept`",
    "`prior_aux`",
    "`...`"
)

description <- c(
    paste0("An object of class `\"formula\"` which determines the linear predictor for the ",
    "ascertainment rate for a particular observation vector. The left hand side must take ",
    "the form `obs_col(group_col, date_col)`, which is syntactic sugar for the observation ",
    "in a given country on a given date. `obs_col` refers to the column of observations in ",
    "the dataframe passed as the `data` argument to `epim()` which is to be modeled. `group_col` ",
    "and `date_col` must be the same as those used in the call to `epirt()`."),
    paste0("A numeric (simplex) vector defining the probability mass function $\\pi_k$ of the ",
    "time distribution from infection to observation."),
    paste0("A string representing the family of the sampling distribution $p(y_t,\\phi)$. Can be ",
    "one of `\"poisson\"`, `\"neg_binom\"`, `\"quasi_poisson\"`, `\"normal\"` or `\"log_normal\"`."),
    paste0("A string representing the link function used to transform the linear predictor. Can be ",
    "one of `\"logit\"`, `\"probit\"`, `\"cauchit\"`, `\"cloglog\"`, `\"identity\"`. Defaults to `\"logit\"`."),
    "same as in `epirt()`, described above.",
    paste0("The prior distribution for the auxiliary parameter $\\phi$, if it exists. ",
    "Only used if family is `\"neg_binom\"` (reciprocal dispersion), `\"quasi_poisson\"` (dispersion), ",
    "`\"normal\"` (standard deviation) or `\"log_normal\"` (sigma parameter)."),
    "Additional arguments for `model.frame`."
)
df <- data.frame(Argument = arguments, Description = description)
knitr::kable(df, caption = "Formal arguments for `epiobs()`", booktabs=TRUE)
```

