---
title: "Partial Pooling"
output:
  bookdown::html_document2:
    theme: flatly
link-citations: yes
bibliography: ../inst/REFERENCES.bib
pkgdown:
  as_is: true
---

<style>
.contents .page-header {
    display: none;
} 
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

<div class="article_container">

# Partial Pooling in epidemia {#sec:partial_pooling}
<hr>

We describe how to partially pool parameters underlying the reproduction 
numbers. This is done using a special operator in the formula passed to 
`epirt()`. If you have previously used any of the **lme4**, 
**nlmer**, **gamm4**, **glmer** or **rstanarm** packages then this 
syntax will be familiar.

A general **R** formula is written as `y ~ model`, where `y` is the response that 
is modeled as some function of the linear predictor which is symbolically 
represented by `model`. `model` is made up of a series of terms separated by 
`+`. In **epidemia**, as in many other packages, parameters can be partially pooled 
by using terms of the form `(expr | factor)`, where both `expr` and `factor` are 
**R** expressions. `expr` is a standard linear model (i.e. treated the same as 
`model`), and is parsed to produce a model matrix. The syntax `(expr | factor)` 
makes explicit that columns in this model matrix have separate effects for 
different levels of the factor variable. 


Of course, separate effects can also be specified using the standard interaction 
operator `:`. This however corresponds to *no pooling*, in that parameters 
at different levels are given separate priors. The `|` operator, on the other 
hand, ensures that effects for different levels are given a common prior. This 
common prior itself has parameters which are given hyperpriors. This allows 
information to be shared between different levels of the factor. To be concrete, suppose 
that the model matrix parsed from `expr` has $p$ columns, 
and that `factor` has $L$ levels. The $p$-dimensional parameter vector for 
the $l$^th^ group can be denoted by $\theta_l$. In **epidemia**, this vector 
is modeled as multivariate normal with an unknown covariance matrix. Specifically,
\begin{equation}
  \theta_{l} \sim N(0, \Sigma),
\end{equation}
where the covariance $\Sigma$ is given a prior. **epidemia** offers the same 
priors for covariance matrices as **rstanarm**; in particular the `decov()` and `lkj()` priors from **rstanarm** can be used. Note that $\Sigma$ is not assumed diagonal, i.e. 
the effects within each level may be correlated.

If independence is desired for parameters in $\theta_l$, we can simply replace 
`(expr | factor)` with `(expr || factor)`. This latter term effectively 
expands into $p$ terms of the form `(expr_1 | factor)`, $\ldots$, `(expr_p | factor)`,
where `expr_1` produces the first column of the model matrix given by `expr`, 
and so on. From the above discussion, the effects are independent across terms, 
and essentially $\Sigma$ is replaced by $p$ one-dimensional covariance matrices 
(i.e. variances).


## Example Formulas
<hr>

The easiest way to become familiar with how the `|` operator works is to 
see a multitude of examples. Here, we give many examples, their interpretations, 
and where possible we compare the models to the no pooling and full pooling equivalents.
For a comprehensive reference on mixed model formulas, please see @Bates_2015.

There are many possible ways to specify intercepts. Table \@ref(tab:intercept-specs) demonstrates some of these, including fully pooled, partially pooled and unpooled. 
Effects may also be partially pooled. This is shown in Table \@ref(tab:cov-specs).

<div class="table">
```{r intercept-specs, echo=FALSE, warning = FALSE, message = FALSE}
library(dplyr)
library(kableExtra)

arguments <- c(
  "`1 + ...`",
  "`region + ...`",
  "`(1 | region) + ...`",
  "`(1 | continent) + ...`"
)

description <- c(
  "Full pooling, common intercept for all regions.",
  "Separate intercepts for each region, not pooled.",
  "Separate intercepts for each region which are partially pooled.",
  "Separate intercepts based on a factor other than `region`, partially pooled."
)

df <- data.frame(Argument = arguments, Description = description)
colnames(df) <- c("Formula R.H.S.", "Interpretation")
knitr::kable(df, caption = "Different intercept specifications. The intercept often has an interpretation as setting $R_0$ in each region. The left hand side of each formula is assumed to take the form `R(region, date)`.", booktabs=TRUE, escape=FALSE) %>%
  kable_paper("hover") %>%
  column_spec(1, bold = T, width = "25em") %>%
  column_spec(2, width = "40em") 
```
</div>

<div class="table">
```{r cov-specs, echo=FALSE}
arguments <- c(
  "`1 + npi + ...`",
  "`1 + npi:region + ...`",
  "`1 + (0 + npi|region) + ...`",
  "`1 + (npi|region) + ...`"
)

description <- c(
  "Full pooling. Effect of NPI the same across all regions.",
  "No pooling. Separate effect in each region.",
  "Partial pooling. Separate effects in each region.",
  "Right hand side expands to `1 + (1 + npi|region)`, and so both the intercept and effect are partially pooled."
)

df <- data.frame(Argument = arguments, Description = description)
colnames(df) <- c("Formula R.H.S.", "Interpretation")
knitr::kable(df, caption = "Different covariate specifications. Here NPI refers to some non-pharamceutical intervention. The left hand side of each formula is assumed to take the form `R(region, date)`.", booktabs=TRUE, escape=FALSE) %>%
  kable_paper("hover") %>%
  column_spec(1, bold = T, width = "25em") %>%
  column_spec(2, width = "40em") 
```
</div>

The final example in Table \@ref(tab:cov-specs) shows that it is important to remember that to parse the term `(expr | factor)`, `epim()` first parses `expr` into a model matrix in the same way as functions like `lm()` and `glm()` parse models. In this case, the intercept term is implicit. Therefore, if this is to be avoided, we must 
explicitly use either `(0 + npi | region)` or `(-1 + npi | region)`.

### Independent Effects

By default, the vector of partially pooled intercepts and slopes for each region 
are correlated. The `||` operator can be used to specify independence. For 
example, consider a formula of the form
```{r eval = FALSE}
R(region, date) ~ npi + (npi || region) + ...
```
The right hand side expands to `1 + npi + (1 | region) + (npi | region) + ...`. Separate intercepts and effects for each region which are partially pooled. The intercept and NPI effect 
are assumed independent within regions.

### Nested Groupings

Often groupings that are nested. For example, suppose we wish to model an 
epidemic at quite a fine scale, say at the level of local districts.
Often there will be little data for 
any given district, and so no pooling will give highly variable estimates of 
reproduction numbers. Nonetheless, pooling at a broad scale, say at the country 
level may hide region specific variations.

If we have another variable, say `county`, which denotes the county to which 
each district belongs, we can in theory use a formula of the form
```{r eval = FALSE}
R(district, date) ~ (1 | county / district) + ...
```
The right hand side expands to `(1 | county) + (1 | county:district)`. 
There is a county level intercept, which is partially 
pooled across different counties. There are also district intercepts 
which are partially pooled *within* each county.

# References
<hr>

</div>