---
title: "Plotting with epidemia"
output:
  bookdown::html_document2:
    theme: cerulean
link-citations: yes
pkgdown:
  as_is: true
fig_width: 9 
fig_height: 6
bibliography: ../inst/REFERENCES.bib
vignette: |
  %\VignetteIndexEntry{Plotting with epidemia} 
  %\VignetteEngine{knitr::rmarkdown} 
  \usepackage[utf8]{inputenc}
---

This vignette demonstrates some of the plotting utilities available in **epidemia**.

**epidemia** provides three plotting functions:
1. `plot_rt`: for plotting the reproduction number
2. `plot_obs`: for plotting the predictve posterior for a single type of observation (for example, daily cases or daily deaths).
3. `plot_infections`: for plotting the number of infections.

The first argument of all three functions is an `epimodel` object returned by calling `epim`

# Fitting a model

We start by fitting a model, the results of which we are going to plot. The model is fitted for three countries: Italy, Austria and Germany.

```{r message=FALSE, cache=T}
library(epidemia)
data("EuropeCovid")
options(mc.cores = parallel::detectCores())

# collect arguments for 'epim'
args <- EuropeCovid
args$algorithm <- "sampling"
args$sampling_args <- list(iter=1e3,control=list(adapt_delta=0.95,max_treedepth=15),seed=12345)
args$group_subset <- c("Italy", "Austria", "Germany")
args$rt <- epirt(
  formula = R(country, date) ~ 1 + lockdown,
  prior = rstanarm::normal(location=0, scale=.5),
  prior_intercept = rstanarm::normal(location=0, scale=2) 
)
fit <- do.call("epim", args)
```

# Default plotting behaviour

## Reproduction number

Calling `plot_rt` with no arguments other than the fitted `epimodel` plots the repdoduction for every group for the entire date range for which the model was fitted.

```{r fig.width=9, fig.height=4, fig.align="center"}
plot_rt(fit)
```

## Observations

To plot observations, the name of the observation type must also be specified using the `type` argument.

```{r plotItaly, fig.width=9, fig.height=4, fig.align="center"}
plot_obs(fit, type="deaths")
```

## Number of infections

```{r fig.width=9, fig.height=4, fig.align="center"}
plot_infections(fit)
```

# Plotting a subset of the groups

Use the `groups` argument to plot a subset of the available group. For example, we can plot the repduction number for just Austria and Germany.

```{r fig.width=9, fig.height=4, fig.align="center"}
plot_rt(fit, groups=c("Austria", "Germany"))
```

# Restricting the date range

We can also subset the date range plotted. For example, we may want to focus on the dates where interventions occur. This is done using the `dates` argument, which is passed a length-2 vector with `c(start_date, end_date)`. Both elements of `dates` must be coercible to `Date` - if the coercion fails for any reason then the whole date range is plotted. 

Other important arguments related to the dates are:

* `date_breaks`: the frequency of the tick labels on the date axis. Default is "2 weeks". This argument is passed to the `date_breaks` argument of `ggplot2::scale_x_date`. See [its documentation](https://ggplot2.tidyverse.org/reference/scale_date.html) for allowed values.
* `date_format`: passed as the the `format` argument to `as.Date` when coercing `dates` to class `Date`. Default is "%Y-%m-%d". This date format should match the date format in the data used to fit the model.

```{r fig.width=9, fig.height=4, fig.align="center"}
plot_rt(fit, dates=c("2020-03-09", "2020-04-06"), date_breaks="1 week")
```

The same plot with a different date format:

```{r fig.width=9, fig.height=4, fig.align="center"}
plot_rt(fit, dates=c("09-03-20", "06-04-20"), date_format="%d-%m-%y", date_breaks="1 week")
```

To restrict the dates from one direction only, set the appropriate element of `dates` to `NA`. 

```{r fig.width=9, fig.height=4, fig.align="center"}
plot_rt(fit, dates=c("2020-03-09", NA))
```


```{r fig.width=9, fig.height=4, fig.align="center"}
plot_rt(fit, dates=c(NA, "2020-04-06"))
```

# Plotting cumulative quantities

For obsersevations and infections set `cumulative=TRUE` to plot the cumulative quantitity.

```{r fig.width=9, fig.height=4, fig.align="center"}
plot_obs(fit, type="deaths", cumulative=TRUE)
```

```{r fig.width=9, fig.height=4, fig.align="center"}
plot_infections(fit, cumulative=TRUE)
```

# Smoothing the repdoduction number

The reproduction number can be smoothed by taking the mean value over a rolling window. The size of this window is controlled using the `smooth` argument to `plot_rt`. It is an integer giving the size of the window in days. The default value of 1 corresponds to no smoothing.

The smoothing is applied separately to each posterior draw.

```{r fig.width=9, fig.height=4, fig.align="center"}
plot_rt(fit, smooth=7)
```
