---
title: "Using Prior Distributions"
output:
  bookdown::html_document2:
    theme: cerulean
link-citations: yes
pkgdown:
  as_is: true
fig_width: 9 
fig_height: 6
bibliography: ../inst/REFERENCES.bib
vignette: |
  %\VignetteIndexEntry{Using prior distributions} 
  %\VignetteEngine{knitr::rmarkdown} 
  \usepackage[utf8]{inputenc}
---

*Please read the [introductory vignette](introduction.html) before this document.* 

# Overview

Recall the parameterisation for the time-varying reproduction number, given by \@ref(eq:rt).
\begin{equation}
R = 2A f\left(X\beta + Zb\right),
(\#eq:rt)
\end{equation}
where $f$ is the standard logistic function. Prior distributions should be specified for the parameters in \@ref(eq:rt), and for additional model parameters. Here we focus on those arising from \@ref(eq:rt).

*Please note that the current default priors may not be appropriate, 
and should not be relied upon. In future versions of the package, 
we hope to set reasonable default priors.*

The priors for $\beta$ are set using the `prior` and 
`prior_intercept` arguments. These will only have an effect if the formula 
specifies covariates and an intercept respectively. These function  similarly 
to the same arguments in [`stan_lm`](https://mc-stan.org/rstanarm/reference/stan_lm.html). 
To understand which families can be used please refer to 
[rstanarm's priors](https://mc-stan.org/rstanarm/reference/priors.html). 


Suppose we are modeling a single group and \@ref(eq:rt) specifies an intercept. If `center=FALSE` in a call to `epim`, then `prior_intercept` sets the prior on the raw intercept. If all other covariates are zero at the beginning of the epidemic then 
this can be viewed as specifying a prior on the initial reproduction rate. We do not recommend 
using `center=TRUE`, because `prior_intercept` then refers to 
the prior on the intercept after all covariates have been centered. This is a difficult 
quantity to intuit for these models.

# Autoscaling `prior`

The scale of the `prior` argument may be adjusted internally to attempt to make the prior 
is weakly informative. This functionality mirrors that used in **rstanarm**. This rescaling can occur both when the default argument is used, and when 
it is user-specified. This adjustment can easily be disabled by setting `autoscale=FALSE` in the 
call to the prior functions. An example of such a call would be `rstanarm::normal(autoscale=FALSE)`.

The automatic scaling rules used for the `prior` is the same as that in **rstanarm**. For clarity, 
we repeat these rules here. Please also refer to [rstanarm's priors](https://mc-stan.org/rstanarm/reference/priors.html) for more information on this.

* If a predictor has only one unique value, no rescaling occurs.
* If it has two unique values, the original scale is divided by the range of the values.
* For more than two unique values, the original scale is divided by the standard deviation of the predictor.

If you are unsure whether rescaling has occurred, call `prior_summary` on a fitted model object.
If an adjustment has been made, this will show both the originally specified prior and the scaled version An example is given below.

```{r priorAS, message=FALSE, warning=FALSE}
library(epidemia)
options(mc.cores = parallel::detectCores())

data("EuropeCovid")
args <- EuropeCovid
args$group_subset <- "France"
args$algorithm <- "meanfield"
args$formula <- R(country,date) ~ 1 + schools_universities + lockdown 
fm <- do.call("epim", args)

# default prior scale not adjusted
prior_summary(fm)
```
Notice that no rescaling occurs and the default priors are used.


```{r priorAS2, message=FALSE, warning=FALSE}
# rescale lockdown indicator
args$data$lockdown <- 100 * args$data$lockdown
fm2 <- do.call("epim", args)
# notice the rescaled prior for lockdown
prior_summary(fm2)

```
Rescaling the lockdown indicator has lead to an adjusted prior on the lockdown coefficient.

# Additional Priors

**epidemia** offers additional priors on top of those in **rstanarm**. Currently there is just one, `shifted_gamma`, which can only be used for the `prior` argument. The addition of this prior is motivated by the priors used in @Flaxman2020. This essentially allows for 
\begin{equation}
\beta_i \sim \text{Gamma}(\alpha,\theta) - \eta,
\end{equation}
where $\alpha$ and $\theta$ are shape and scale parameters, and $\eta$ is a shift to allow for priors with support below zero.

## Priors on Covariance Matrices

Consider a single 'random-effects' term in the formula. A good example would be `(1 + lockdown | country)`. For each group the intercepts and slopes are treated as zero-mean multivariate normal with unknown covariance matrix $\Sigma$. For our example term, fix one level implied by the factor `country`, say `"Italy"`. For this group, there are two coefficients - a random intercept for Italy, and a random slope. Therefore $\Sigma$ has dimension $2 \times 2$. Coefficients for different groups are treated as independent. 

**epidemia** borrows from **rstanarm** and uses the [`decov`](rstanarm priors) prior for $\Sigma$. This decomposes $\Sigma$ into a variance vector and a correlation matrix. An LKJ prior is used on the correlation matrix, while the variance is decomposed into the product of a simplex vector and the trace of $\Sigma$. The simplex vector is given a symmetric Dirichlet prior, which the trace is once again decomposed as $tr(\Sigma) = J \kappa$, where $J$ is the order of $\Sigma$. Finally $\kappa$ is assigned a scale-invariant prior - specifically a Gamma distribution with give shape and scale hyperparameters. For more information on this, please see [rstanarm priors](http://mc-stan.org/rstanarm/reference/priors.html) and this [vignette](http://mc-stan.org/rstanarm/articles/glmer.html).

When the left hand side of the formula has only one term (for example `(1 | country)`), this prior on the covariance simplifies considerably. $\Sigma$ simply reduces to $\kappa$, which has a gamma prior.

# References
