---
title: "Resolving Problems"
output: 
  github_document:
    fig_width: 10
    fig_height: 7
vignette: >
  %\VignetteIndexEntry{Resolving Problems}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

## SARS 2003  - using cumulatives to obtain starting values for MCMC

We use the SARS data from 2003 as an example. 

In the example below, the sampler would have difficulties converging with the
default random starting values of rstan. It usually stays in a part of the parameter space where all of the population gets infected. 

Below is one approach how to resolve this. This is at the moment
hard-coded. Future versions of epidemia may provide a more convenient
way of doing this.

We first fit the model to cumulative counts to get starting values for
the main run (the fit to cumulative counts will not be correct itself
due an assumed independence of the cumulative counts).  

First we load the data.

```{r}
library(epidemia)
data("SARS2003")
options(mc.cores = parallel::detectCores())
print(SARS2003)
```
Then we set up and run the model for cumulative observed cases.
```{r SARS2003}
sars <- SARS2003
sars$incidence <- c(rep(NA,20),cumsum(sars$incidence)) ## pad before initialisation
sars$sarsdate <- as.Date("2003-01-01")+seq(0,along.with=sars$incidence)
obs <- !is.na(sars$incidence)

args <- list(formula=Rt(country,date)~rw(date,3),
     data=data.frame(country="A",date=sars$sarsdate),
             obs=list(
                 incidence=list(
                     odata=data.frame(country="A",
                                      date=sars$sarsdate[obs],incidence=sars$incidence[obs]),
                     rates=list(means=data.frame(factor("A"),1),
                                scale=.01),
                     pvec=c(.25,.5,.75,1),
                     ptype="distribution"
                 )
             ),
             seed_days=7,
             algorithm="sampling",
             r0=3,
             pops=data.frame(country="A",pop=1e6),
             si=sars$si,
             prior = rstanarm::normal(location=0,scale=.2),
             prior_intercept = rstanarm::normal(location=0,scale=.5),
             prior_tau = rstanarm::exponential(rate=4)             
     )
```

```{r SARS_cumulative_pre, results='hide', cache=TRUE}
args$debug=TRUE ## to get original parameter values
args$sampling_args <- list(iter=100,control=list(adapt_delta=0.95,max_treedepth=15),seed=77239,chains=1)
fitpre <- do.call("epim",args)
```
The above run may give some error messages, but they can be ignored as this run is only used to get starting values for the chains of the main run.

Now update the model to work with individually reported cases.
```{r SARS_prepdata}
sars <- SARS2003
sars$incidence <- c(rep(NA,20),sars$incidence) ## pad before initialisation
sars$sarsdate <- as.Date("2003-01-01")+seq(0,along.with=sars$incidence)
obs <- !is.na(sars$incidence)
args$obs=list(
    incidence=list(
        odata=data.frame(country="A",
                         date=sars$sarsdate[obs],incidence=sars$incidence[obs]),
        rates=list(means=data.frame(factor("A"),1),
                   scale=.01),
        pvec=c(.25,.25,.25,.25),
        ptype="density"
    )
)
```

Next we prepare the initialisation values. This is very crude at the moment and better ways probably do exist. 
```{r}
initf <- function(){
    i <- sample(1:50,1)
    res <- lapply(rstan::extract(fitpre$stanfit),
                  function(x) {
                      if (length(dim(x))==1){
                          as.array(x[i])
                      }
                      else if (length(dim(x))==2)
                          x[i,]
                      else x[i,,]
                  }
                  )
    for (j in names(res)){
        if (length(res[j])==1)
            res[[j]] <- as.array(res[[j]])
    }
    res$tau_raw <- c(res$tau_raw)
    res$noise<- NULL
    res
}
args$sampling_args <- list(iter=1000,control=list(adapt_delta=0.95,max_treedepth=15),seed=713,init=initf)
```


Now we starte the main sampling run.

```{r SARS_sampling, cache=TRUE, results='hide'}
fit <- do.call("epim",args)
```


And here are the resulting fits. 
```{r SARScumulativeplot}
library(gridExtra)
grid.arrange(plot_rt(fit),
             plot_obs(fit,"incidence"),
             nrow=2)
```
